Notes on Big MRA

### 2018-09-12 Notes/notes_Will.tex
Editing

@@

@@

>>>>

\newcommand{\R}{\mathbb{R}}

@@

\renewcommand{\P}{\mathbb{P}}

@@

\section{Impossibility of detection}

Let's consider an even easier problem, in which someone hands us intervals of length $L$ that either contain the full signal plus noise, or contain just noise, and we are asked to determine which intervals contain signal. Let's assume too that we know the signal $x$, that we know the probability $q$ that one of the intervals contains a signal, and that we know the noise variance $\sigma$. Since we have the full generative model, we can always generate more data as needed; therefore it's enough to consider the hardness of deciding whether a single interval contains signal or not.

Let's abstract this problem. We have two known vectors $\theta_0$ and $\theta_1$ in $\R^L$. There is a random variable $\ep$ taking values 0 or 1 with probabilities $q$ and $1-q$, respectively. We observe a random vector $X \in \R^L$ with the following distribution: $X | (\ep = i) \sim N(\theta_i,\sigma^2)$; in other words, if $\ep = 0$, then $X \sim N(\theta_0,\sigma^2)$, and if $\ep = 1$, then $X \sim N(\theta_1,\sigma^2)$.

We observe $X$, and our task is to predict $\ep$. How well can we do it? First, let's assume without loss of generality that $q \ge 1-q$. Then if we make the constant prediction $\hat{\ep} \equiv 0$, we will be correct with probability $q$. So the question is, can we do better than this? We will prove that as $\sigma \to \infty$, the answer is no. More precisely:

\begin{prop}
Let $\hat{\ep}$ be any predictor of $\ep$. Then 
%
\begin{align}
%
\lim_{\sigma \to \infty} \P[ \hat{\ep} = \ep ] \le q.
%
\end{align}

\end{prop}

To begin the proof, we use a variant of the Neyman-Pearson Lemma to derive the best (deterministic) predictor $\hat{\ep}$. We take any predictor $\hat{\ep}$, and we let $S$ be the set of $X$'s where $\hat{\ep} = 1$. Then the probability that $\hat{\ep}$ fails is:
%
\begin{align}
\label{eq:neyman-pearson}
%
\P[\hat{\ep} \ne \ep] 
&= q \P_0[\hat{\ep} = 1 ] + (1-q) \P_1[\hat{\ep} = 0]
    \nonumber \\
&= q \P_0[\hat{\ep} = 1] + (1-q) (1 - \P_1[\hat{\ep} = 1])
    \nonumber \\
&= q \P_0[\hat{\ep} = 1] + (1-q) - (1-q) \P_1[\hat{\ep} = 1]
    \nonumber \\
&= (1-q)  + \int_S (q f_0(x) - (1-q) f_1(x))dx,
%
\end{align}
%
where $f_i(x)$ is the normal density with mean $\theta_i$ and variance $\sigma^2$, and the notation $\P_i$ means probability conditional on the event $\ep = \theta_i$; that is, $\P_i[A] = \P[A | \ep=i]$. The integral \eqref{eq:neyman-pearson} is minimized by taking $S$ to be the set:
%
\begin{align}
%
S = \left\{x : \frac{f_0(x)}{f_1(x)} \le \frac{1-q}{q} \right\}.
%
\end{align}
%
In other words, if $\Lambda(x) = f_0(x) / f_1(x)$ and $b = (1-q) / q$, then the best predictor of $\ep$ based on $X$ is:
%
\begin{align}
%
\ep = 
\begin{cases}
1,  \text{ if } \Lambda(x) \le b \\
0,  \text{ if } \Lambda(x) \ge b
\end{cases}.
%
\end{align}

Taking logarithms, the set $S$ can be rewritten as the set of $x$ where:
\begin{align}
%
-\| x - \theta_0\|^2 \le - \|x - \theta_1\|^2 + 2 \sigma^2 \log(b),
%
\end{align}
%
or equivalently
%
\begin{align}
%
\langle x , \theta_1 - \theta_0 \rangle 
    \ge \frac{\|\theta_1\|^2 - \|\theta_0\|^2}{2} - \sigma^2 \log(b).
%
\end{align}
%

Now let's compute the probability of failure conditional on the event $\ep = 0$. In this case, failure occurs when $X \in S$. Since $X|(\ep=0) \sim N(\theta_0,\sigma^2)$, we can write $X | (\ep=0) = \sigma Z + \theta_0$, where $Z \sim N(0,1)$. Consequently,
%
\begin{align}
%
\langle X , \theta_1 - \theta_0 \rangle 
&= \sigma \langle Z , \theta_1 - \theta_0 \rangle 
    + \langle \theta_0,\theta_1 - \theta_0 \rangle
    \nonumber \\
&= \sigma \langle Z , \theta_1 - \theta_0 \rangle 
            + \langle \theta_0,\theta_1 \rangle - \|\theta_0\|^2
%
\end{align}
%
and failure occurs when 
%
\begin{align}
%
\sigma\langle Z, \theta_1 - \theta_0 \rangle 
&\ge \frac{\|\theta_1\|^2 + \|\theta_0\|^2}{2} - \langle \theta_0,\theta_1 \rangle
    - \sigma^2 \log(b)
    \nonumber \\
&= \frac{1}{2} \|\theta_1 - \theta_0\|^2 - \sigma^2 \log(b)
%
\end{align}
%
or equivalently
%
\begin{align}
%
Y \ge \frac{c}{\sigma} - \sigma \log(b)
%
\end{align}
%
where $c = \|\theta_1 - \theta_0\|^2 / 2$ and $Y = \langle Z , \theta_1 - \theta_0 \rangle \sim N(0,\|\theta_1 - \theta_0\|^2)$. Just for simplicity let's assume $\| \theta_1 - \theta_0\| = 1$, so that $Y \sim N(0,1)$. So:
%
\begin{align}
%
\P_0[\hat{\ep} = 1] = \P\left[Y \ge \frac{c}{\sigma} - \sigma^2 \log(b) \right],
    \quad Y \sim N(0,1).
%
\end{align}
%
Similarly,
%
\begin{align}
%
\P_1[\hat{\ep} = 0] = \P\left[Y \ge \frac{c}{\sigma} + \sigma^2 \log(b) \right],
    \quad Y \sim N(0,1).
%
\end{align}
%
So the overall probability of failure is:
%
\begin{align}
%
\P[\hat{\ep} \ne \ep]
= q \P\left[Y \ge \frac{c}{\sigma} - \sigma^2 \log(b) \right]
    + (1-q) \P\left[Y \ge \frac{c}{\sigma} + \sigma^2 \log(b) \right].
%
\end{align}
%
Now, if $q = 1/2$, the $b = 1$ and $\log(b) = 0$. Then the probability of failure is simply:
%
\begin{align}
%
\P\left[Y \ge \frac{c}{\sigma} \right] \longrightarrow \frac{1}{2} = q
    \quad \text{as} \quad \sigma \to \infty.
%
\end{align}

On the other hand, if $q > 1-q$, then $\log(b) < 0$. Then 
%
\begin{align}
%
\P\left[Y \ge \frac{c}{\sigma} - \sigma^2 \log(b) \right] \longrightarrow 0
%
\end{align}
%
while
%
\begin{align}
%
\P\left[Y \ge \frac{c}{\sigma} + \sigma^2 \log(b) \right] \longrightarrow 1
%
\end{align}
%
as $\sigma \to \infty$. Consequently,
%
\begin{align}
%
\P[\hat{\ep} \ne \ep]
&= q \P\left[Y \ge \frac{c}{\sigma} - \sigma^2 \log(b) \right]
    + (1-q) \P\left[Y \ge \frac{c}{\sigma} + \sigma^2 \log(b) \right]
    \nonumber\\
& \longrightarrow 1-q  \quad \text{as} \quad  \sigma \to \infty.
%
\end{align}
%
That is, the probability of success converges to $q$, which completes the proof.
